Evolution of the AGN UV luminosity function

### 2017-09-20 Paper/paper.tex
Update text

@@

The smooth evolution of the luminosity function parameters apparant in
Figure~\ref{fig:evoln} suggests that it should be possible to describe
the quasar UV luminosity function evolution from $z\sim 7$ to $0$
using far fewer number of parameters.  Such descriptions have been put
forward in the literature for the X-ray and bolometric luminosity
functions \citep[e.g.,][]{2007ApJ...654..731H, 2015MNRAS.451.1892A}.
Such ``global'' descriptions of the luminosity function evolution are
useful as they give a continuous description of the luminosity
function.  This makes them valuable while developing an understanding
of the physics behind the luminosity function. We therefore obtain
estimates of the luminosity function in which the evolution of the
four parameters of the double power law models in
Equation~(\ref{eqn:dpl}) is parameterised to write
\begin{align}
  &\phi_*(z) = f_0(\{c_{0,i}\}, z)\nonumber\\
  &M_*(z) = f_1(\{c_{1,i}\}, z)\nonumber\\
  &\alpha(z) = f_2(\{c_{2,i}\}, z)\nonumber\\
  &\beta(z) = f_3(\{c_{3,i}\}, z),
\end{align}
where the $c_i$'s are the new model parameters.  The joint posterior
probability distribution of these parameters can be now written as
\begin{equation}
  p(\{c_i\} | \{M_i, z_i\}) \propto p(\{c_i\})p(\{M_i, z_i\} | \{c_i\}),
\end{equation}
where the likelihood 
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \{c_i\}),
\end{equation}
is now given by $\phi(M,z)$ with suitable normalisation.  Note that
$\phi(M,z)$ is simply 
\begin{multline}
  \phi(M,z) = \phi_*(z) \left[10^{0.4(\alpha(z)+1)(M-M_*(z))}\right. \\ \left.+ 10^{0.4(\beta(z)+1)(M-M_*(z))}\right]^{-1}.
\end{multline}
The ngative logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ is
straightforward generalisation of Equation~(\ref{eqn:S}), so that
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM\int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz},
  \label{eqn:S2}
\end{multline}
where now $N$ is the total number of quasars included in the analysis,
and the integrals in the second term on the right-hand side are over
the complete surveyed range of luminosity and redshift.

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.6\textwidth,keepaspectratio]{rhoqso_allz.pdf}
  \end{center}
  \caption{AGN number density evolution, showing the systematic offset in BOSS quasars.}
  \label{fig:rhoqso_allz}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.6\textwidth,keepaspectratio]{rhoqso.pdf}
  \end{center}
  \caption{AGN number density evolution with some BOSS quasars
    removed.}
  \label{fig:rhoqso}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{evolution_seldata.pdf}
  \end{center}
  \caption{Parameter evolution with some BOSS quasars removed.  We (1)
    use \emph{only} BOSS quasars for $z=2.2$--$3.5$, and (2) use only
    $p>0.9$ quasars in the $z>3.7$ sample of Richards.}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth,keepaspectratio]{mosaic_withGlobal.pdf}
  \end{center}
  \caption{Individual and composite luminosity function fits for
    $z=0.3$ to $2.6$.  They do not agree for $2.2 < z < 2.6$ because
    composite fit does not consider quasars at those redshifts.}
  \label{fig:mosaic_withGlobal}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth,keepaspectratio]{mosaic2_withGlobal.pdf}
  \end{center}
  \caption{Individual and composite luminosity function fits for
    $z=2.6$ to $6.5$.  They do not agree for $2.6 < z < 2.8$ because
    composite fit does not consider quasars at those redshifts.}
  \label{fig:mosaic2_withGlobal}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{evolution_global.pdf}
  \end{center}
  \caption{Luminosity function parameter evolution in the global model.}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.6\textwidth,keepaspectratio]{rhoqso_withGlobal.pdf}
  \end{center}
  \caption{AGN number density evolution in the global model.}
  \label{fig:rhoqso}
\end{figure*}

>>>>

\begin{figure*}
  \begin{center}
    \includegraphics[height=0.88\textheight,keepaspectratio]{mosaic_small_global.pdf}
  \end{center}
  \caption{Luminosity function estimates from $z=0.3$ to $2.6$.
    Similar to Figure~\ref{fig:mosaic}, the symbols show our inferred
    binned luminosity functions.  In each redshift bin, the yellow
    curves show the best-fit double power law luminosity function in
    that redshift bin.  The green curves show the posterior
    distribution of the global model.  The global model agrees very
    well with the data in all redshift bins, except those spanned by
    BOSS, as BOSS quasars are excluded from the global analysis.}
  \label{fig:mosaic_global}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{evolution_global.pdf}
  \end{center}
  \caption{Luminosity function parameter evolution in the global
    model.  The symbols show the best fit parameters with 68\%
    uncertainties from the double power law fits to the data in
    redshift bins.  These points are identical to those in
    Figure~\ref{fig:evoln}, except that bins with BOSS qsos ($z\sim
    2$--$3$) are not shown.  In each panel, the black solid curve
    shows the best fit global model, and the coloured curves show a
    random sample drawn from the global posterior.}
  \label{fig:evoln_global}
\end{figure*}

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth,keepaspectratio]{rhoqso_withGlobal.pdf}
  \end{center}
  \caption{AGN number density evolution in the global model.}
  \label{fig:rhoqso}
\end{figure}

@@

The smooth evolution of the luminosity function parameters apparant in
Figure~\ref{fig:evoln} suggests that it should be possible to describe
the quasar UV luminosity function evolution from $z\sim 7$ to $0$
using far fewer number of parameters.  Such descriptions have been put
forward in the literature for the X-ray and bolometric luminosity
functions \citep[e.g.,][]{2007ApJ...654..731H, 2015MNRAS.451.1892A}.
Such ``global'' descriptions of the luminosity function evolution are
useful as they give a continuous description of the luminosity
function.  This makes them valuable while developing an understanding
of the physics behind the luminosity function. We therefore obtain
estimates of the luminosity function in which the evolution of the
four parameters of the double power law models in
Equation~(\ref{eqn:dpl}) is parameterised to write
\begin{align}
  &\phi_*(z) = f_0(\{c_{0,i}\}, z)\nonumber\\
  &M_*(z) = f_1(\{c_{1,i}\}, z)\nonumber\\
  &\alpha(z) = f_2(\{c_{2,i}\}, z)\nonumber\\
  &\beta(z) = f_3(\{c_{3,i}\}, z),
\label{eqn:global}
\end{align}
where the $c_i$'s are the new model parameters.  The joint posterior
probability distribution of these parameters can be now written as
\begin{equation}
  p(\{c_i\} | \{M_i, z_i\}) \propto p(\{c_i\})p(\{M_i, z_i\} | \{c_i\}),
\end{equation}
where the likelihood 
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \{c_i\}),
\end{equation}
is now given by $\phi(M,z)$ with suitable normalisation.  Note that
$\phi(M,z)$ is simply 
\begin{multline}
  \phi(M,z) = \phi_*(z) \left[10^{0.4(\alpha(z)+1)(M-M_*(z))}\right. \\ \left.+ 10^{0.4(\beta(z)+1)(M-M_*(z))}\right]^{-1}.
\end{multline}
The ngative logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ is
straightforward generalisation of Equation~(\ref{eqn:S}), so that
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM\int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz},
  \label{eqn:S2}
\end{multline}
where now $N$ is the total number of quasars included in the analysis,
and the integrals in the second term on the right-hand side are over
the complete surveyed range of luminosity and redshift.

We consider models in which the evolution of the four double power law
parameter is modelled independently as in
Equations~(\ref{eqn:global}).  This approach has been termed as
``flexible double power law'' by \citet{2015MNRAS.451.1892A}.  For the
parameters $\phi_*$, $M_*$ and $\alpha$ we assume that the functions
$f_i$s in Equations~(\ref{eqn:global}) are Chebyshev polynomials in
$1+z$.  Thus the functions $f_i$s are written as
\begin{equation}
  f_i=\sum_{j=0}^{n_i}c_{i,j}T_j(1+z),
  \label{eqn:cbs}
\end{equation}
where $c_{i,j}$ are the parameters that appear in
Equations~(\ref{eqn:global}) and $T_j(1+z)$ are Chebyshev polynomials
of the first kind.  We try successively higher orders of Chebyshev
polynomials in order to arrive at a good fit with the data.  We find
that $\phi_*$, $M_*$ and $\alpha$ prefer quadratic, cubic, and linear
evolutions in $1+z$.

The faint-end slope $\beta$ seems to require a break in its evolution
at $z\sim 3$, as discussed in the previous section.  As a result, it
is difficult to model the evolution of $\beta$ using the Chebyshev
expansion of Equation~(\ref{eqn:cbs}).  Instead, we consider a double
power law model for $\beta$ and write
\begin{equation}
  f_3(1+z)=h+\frac{f_0}{10^{a*\zeta}+10^{b*\zeta}},
\end{equation}
where
\begin{equation}
  \zeta = \log_{10}\left(\frac{1+z}{1+z_0}\right),
\end{equation}
thus resulting in a five-parameter model with parameters $h$, $f_0$,
$z_0$, $a$ and $b$.  The parameters $a$ and $b$ thus determine the low
and high redshift slopes of this evolution, with a break at redshift
$z_0$.  This is similar to the model of \citet{2007ApJ...654..731H},
who also favoured a broken power law model for the evolution of the
faint-end slope of the bolometric luminosity function of quasars.

Our resultant 14-parameter global model for the luminosity function
evolution in shown in Figure~\ref{fig:mosaic_global}, in which the
green curves show a random draw from the posterior probability
distibution function.  For comparison, the symbols show the luminosity
function binned in luminosity and redshift, and the yellow curves show
the posterior distribution of the double power law luminosity
functions in various redshift bins, as in Figure~\ref{fig:mosaic}.  We
find that the global model is in excellent agreement with the binned
models.  The agreement is poor for the BOSS quasars as these are not
described by the global model, as discussed in the previous section.

Figure~\ref{fig:evoln_global} shows the parameter evolution in the
global model, by comparing it with the results from the fits in
individual redshift bins shown in Figure~\ref{fig:evoln}.  The
agreement between the two models is very good, showing an evolution
towards a steep faint-end slope of the luminosity function at high
redshifts, a rapid brightening of the break luminosity.  Note that the
redshift bins with BOSS qsos are not shown in
Figure~\ref{fig:evoln_global}, which explains the difference from
Figure~\ref{fig:evoln}.

@@

%% \begin{figure*}
%%   \begin{center}
%%     \includegraphics[width=0.6\textwidth,keepaspectratio]{rhoqso_allz.pdf}
%%   \end{center}
%%   \caption{AGN number density evolution, showing the systematic offset in BOSS quasars.}
%%   \label{fig:rhoqso_allz}
%% \end{figure*}

%% \begin{figure*}
%%   \begin{center}
%%     \includegraphics[width=0.6\textwidth,keepaspectratio]{rhoqso.pdf}
%%   \end{center}
%%   \caption{AGN number density evolution with some BOSS quasars
%%     removed.}
%%   \label{fig:rhoqso}
%% \end{figure*}

%% \begin{figure*}
%%   \begin{center}
%%     \includegraphics[width=\textwidth]{evolution_seldata.pdf}
%%   \end{center}
%%   \caption{Parameter evolution with some BOSS quasars removed.  We (1)
%%     use \emph{only} BOSS quasars for $z=2.2$--$3.5$, and (2) use only
%%     $p>0.9$ quasars in the $z>3.7$ sample of Richards.}
%% \end{figure*}

%% \begin{figure*}
%%   \begin{center}
%%     \includegraphics[width=\textwidth,keepaspectratio]{mosaic_withGlobal.pdf}
%%   \end{center}
%%   \caption{Individual and composite luminosity function fits for
%%     $z=0.3$ to $2.6$.  They do not agree for $2.2 < z < 2.6$ because
%%     composite fit does not consider quasars at those redshifts.}
%%   \label{fig:mosaic_withGlobal}
%% \end{figure*}

%% \begin{figure*}
%%   \begin{center}
%%     \includegraphics[width=\textwidth,keepaspectratio]{mosaic2_withGlobal.pdf}
%%   \end{center}
%%   \caption{Individual and composite luminosity function fits for
%%     $z=2.6$ to $6.5$.  They do not agree for $2.6 < z < 2.8$ because
%%     composite fit does not consider quasars at those redshifts.}
%%   \label{fig:mosaic2_withGlobal}
%% \end{figure*}
